---
title: "Statistical Power"
subtitle: ""
author: "Ted Ellsworth, PhD"
date: today
format:
  beamer:
    theme: default
    aspectratio: 169
    slide-level: 2
    pdf-engine: xelatex
    include-in-header: ../../assets/beamer-header.tex
    keep-tex: true
---


## Free statistical consulting for Yale researchers

\vspace{2.8em}

\begin{center}

\large
Study design $\cdot$ analysis $\cdot$ visualization $\cdot$ writing.

\vspace{1.4em}

\normalsize
Walk-ins, appointments, and workshops --- free for all Yale affiliates.

\vspace{2.8em}

\textcolor{statsecondary}{%
  \large\textbf{library.yale.edu/statlab}%
}

\end{center}


## Introduction

\vspace{1.4em}

\begin{center}
\large A committee, a reviewer, an IRB, or a funder asks: \textbf{how many subjects do you need?}
\end{center}

\vspace{1.6em}

\begin{center}
\small
\setlength{\tabcolsep}{9pt}
\renewcommand{\arraystretch}{1.35}
\begin{tabular}{>{\raggedright\arraybackslash}p{5.9cm}>{\raggedright\arraybackslash}p{6.9cm}}
\toprule
\textbf{What you may have been told} & \textbf{What we will actually do} \\
\midrule
Power is a number you look up once & Power is a property of a design you can change \\
Power is about sample size         & Sample size is one of three levers \\
There is a formula, so use it      & The formula describes one study. Code describes yours. \\
\bottomrule
\end{tabular}
\end{center}


## Schedule

\vspace{0.3em}

\begin{center}
\footnotesize
\setlength{\tabcolsep}{7pt}
\renewcommand{\arraystretch}{1.05}
\begin{tabular}{c>{\raggedright\arraybackslash}p{7.4cm}>{\raggedright\arraybackslash}p{3.1cm}}
\toprule
\textbf{Min} & \textbf{Block} & \textbf{Who is working} \\
\midrule
15 & What power is, and what it looks like                      & Lecture \\
15 & The formula, worked by hand, and the effect it cannot see  & Work \\
15 & \textcolor{statsecondary}{\textbf{Your turn 1:}} a tutoring study, checked in the EGAP app & \textbf{You work} \\
10 & Where the numbers come from: guessing well                 & Discussion \\
\midrule
5  & \textit{Break}                                             & --- \\
\midrule
15 & Power in R: one formula, a grid of assumptions             & Demo \\
15 & \textcolor{statsecondary}{\textbf{Your turn 2:}} build the sweep, find the $N$ you need & \textbf{You code} \\
15 & When the formula stops describing your study: simulation   & Demo \\
5  & What voids your calculation                                & Lecture\\
10 & \textcolor{statsecondary}{\textbf{Your turn 3:}} your own design, on the scaffold & \textbf{You code} \\
\bottomrule
\end{tabular}
\end{center}



## Learning Objectives

\vspace{0.3em}

\small By the end of the session you should be able to:

\vspace{0.4em}

\begin{center}
\scriptsize
\setlength{\tabcolsep}{7pt}
\renewcommand{\arraystretch}{1.12}
\begin{tabular}{>{\raggedright\arraybackslash}p{7.3cm}>{\raggedright\arraybackslash}p{5.4cm}}
\toprule
\textbf{You will be able to} & \textbf{You will show it by} \\
\midrule
\textbf{Name} the three ingredients of power and \textbf{identify} which you control
  & Reading a design and saying which lever is available \\
\textbf{Compute} power for a two-arm difference in means, and the \textbf{smallest effect} a design can see
  & Doing it by hand, then reproducing it in the EGAP calculator \\
\textbf{Write} the formula as an R function and \textbf{sweep} it over a grid of sample sizes and effects
  & Returning the smallest $N$ that reaches 80\% power \\
\textbf{Defend} the effect size you assumed, naming its source
  & Reporting power as a \textit{range}, not one number \\
\textbf{Diagnose} when the closed form no longer describes your study
  & Pointing to the assumption your design breaks \\
\textbf{Simulate} power for a design the formula cannot handle
  & Generating data from your estimand, fitting your estimator, counting rejections \\
\bottomrule
\end{tabular}
\end{center}



## Type I and II Errors

- Most diagnostic statistics seek to identify two types of errors:

  - *Type I Errors* ($\alpha$) refer to the probability that you will claim an effect is *present* when the **null hypothesis is true**.
  - *Type II Errors* ($\beta$) refer to the probability that you will claim an effect is *absent* when the **null hypothesis is false**.

- Power is what is left over from a Type II error: $\text{power} = 1 - \beta$.


## What is Statistical Power? 

- Less formally: power is a study's ability to *distinguish signal from noise.*

  - The *signal* is the real effect.
  - The *noise* is everything else (e.g., sampling variability, measurement error, differences between units, etc.)


## What Power Looks Like

```{r power-picture, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7.6, fig.height=2.75, dev="png", dpi=300, fig.align="center"}
library(ggplot2)
cv <- qnorm(0.975)
zz <- seq(-3.6, 6.2, length.out = 700)
dd <- rbind(
  data.frame(z = zz, y = dnorm(zz, 0),    curve = "If the null is true"),
  data.frame(z = zz, y = dnorm(zz, 2.24), curve = "If the 7-point effect is real"))

ggplot(dd, aes(z, y)) +
  geom_area(data = subset(dd, curve != "If the null is true" & z >= cv),
            fill = "#4A90C2", alpha = 0.55) +
  geom_area(data = subset(dd, curve == "If the null is true" & z >= cv),
            fill = "#B41E1E", alpha = 0.70) +
  geom_line(aes(colour = curve), linewidth = 0.75) +
  geom_vline(xintercept = cv, colour = "grey20", linewidth = 0.4, linetype = "dashed") +
  annotate("text", x = 3.6, y = 0.20, label = "power = 0.61",
           colour = "#0B2E52", size = 3.1, fontface = "bold") +
  annotate("text", x = 2.45, y = 0.045, label = "alpha/2 == 0.025", parse = TRUE,
           colour = "#B41E1E", size = 2.7, hjust = 0) +
  annotate("text", x = cv, y = 0.43, label = "reject to the right of 1.96",
           colour = "grey20", size = 2.7, hjust = -0.04) +
  scale_colour_manual(values = c("If the null is true" = "#B41E1E",
                                 "If the 7-point effect is real" = "#0B2E52"),
                      name = NULL) +
  # 1.96 and 2.24 nearly collide on the axis; drop the second label a row
  scale_x_continuous(breaks = c(0, cv, 2.24), labels = c("0", "1.96", "\n2.24")) +
  labs(x = "Effect, measured in standard errors", y = NULL) +
  theme_minimal(base_size = 9) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.major.x = element_line(linewidth = 0.25, colour = "grey88"),
        axis.text.y = element_blank(),
        axis.title = element_text(colour = "grey30"),
        legend.position = "bottom", legend.margin = margin(t = -6))
```

\vspace{-0.2em}

\begin{center}
\footnotesize \textcolor{errorred}{\textbf{Red}} is the study you would run if there were no effect; the shaded sliver past 1.96 is $\alpha$ --- the false positives you agreed to tolerate. \textcolor{statsecondary}{\textbf{Blue}} is the study you expect if the effect is real; \textbf{the shaded area past the same line is power}.
\end{center}


## Type II Error Tolerance

- When conducting power analysis, our goal is to derive the probability of *detecting* a positive effect when the null hypothesis is false.
- Many studies use $\text{power} \geq 80\%$ (equivalently, $\beta \leq 0.20$) as a threshold:
  - Implies that there is a $20\%$ chance that you will *not* find an effect that is present.
  - Better to treat as a heuristic, rather than an explicit benchmark.


## Core Notation

In order to estimate statistical power, we need four qualities:

\vspace{0.5em}

\begin{center}
\small
\setlength{\tabcolsep}{7pt}
\renewcommand{\arraystretch}{1.35}
\begin{tabular}{c>{\raggedright\arraybackslash}p{3.1cm}>{\raggedright\arraybackslash}p{7.9cm}}
\toprule
\textbf{Symbol} & \textbf{Quality} & \textbf{What it is} \\
\midrule
$N$              & Sample size        & How many subjects you collect in total. \\
$\mu_t - \mu_c$  & Effect size        & The true difference between the treatment and control means: the \textit{signal}. \\
$\sigma$         & Standard deviation & How much outcomes vary from one another: the \textit{noise}. \\
$\alpha$         & Significance level & The Type I error rate you are willing to tolerate; it sets the evidence threshold. \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.4em}

\begin{center}
\small Note we will use $\mu_t - \mu_c$ throughout to denote effect size. Where space is tight --- in code, and in the appendix --- we abbreviate it $\delta$. 
\end{center}


## The Three Ingredients of Statistical Power

\vspace{0.5em}

\begin{center}
\small
\setlength{\tabcolsep}{7pt}
\renewcommand{\arraystretch}{1.35}
\begin{tabular}{>{\raggedright\arraybackslash}p{3.3cm}cc>{\raggedright\arraybackslash}p{6.2cm}}
\toprule
\textbf{Ingredient} & \textbf{Symbol} & \textbf{Control?} & \textbf{What it does} \\
\midrule
Strength of the treatment & $\mu_t - \mu_c$ & Partly & Dose, framing, and duration are yours to set; a stronger treatment is easier to detect. \\
Background noise & $\sigma$ & Partly & Noisier outcomes bury the signal, but better measurement and covariates quiet it. \\
Experimental design & $N$, $\alpha$ & \textbf{Yes} & Sample size, randomization scheme, and covariate adjustment. \\
\bottomrule
\end{tabular}
\end{center}


## Key Formula for Calculating Power

\vspace{0.2em}

\begin{center}
$\displaystyle \text{power} \;=\; \Phi\!\left( \frac{|\mu_t - \mu_c|\sqrt{N}}{2\sigma} \;-\; \Phi^{-1}\!\left(1-\frac{\alpha}{2}\right) \right)$
\end{center}

\vspace{0.8em}

\begin{center}
\small
\setlength{\tabcolsep}{7pt}
\renewcommand{\arraystretch}{1.35}
\begin{tabular}{c>{\raggedright\arraybackslash}p{3.1cm}>{\raggedright\arraybackslash}p{7.9cm}}
\toprule
\textbf{Symbol} & \textbf{Quality} & \textbf{What it is} \\
\midrule
$\Phi$      & Normal CDF         & The CDF of the standard normal distribution. \\
$\Phi^{-1}$ & Inverse normal CDF & Its inverse; returns the critical value for a given tail probability. \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.2em}

\scriptsize Source: Coppock, \textit{10 Things to Know About Statistical Power} (EGAP). EGAP writes this quantity as $\beta$; the dominant convention reserves $\beta$ for the Type II error rate, so that $\text{power} = 1-\beta$. We follow the latter.

## A Running Example: Does the Message Move Opinion?

\begin{notebox}
\small You are fielding a survey experiment with \textbf{164 respondents}, randomly assigned 50/50 to read a short informational message (treatment) or no message (control). Support for the policy is measured on a 0--100 feeling thermometer, and a pilot study supplies your assumptions.
\end{notebox}

\vspace{0.2em}

\begin{center}
\small
\setlength{\tabcolsep}{7pt}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{cc>{\raggedright\arraybackslash}p{7.4cm}}
\toprule
\textbf{Symbol} & \textbf{Value} & \textbf{Where it comes from} \\
\midrule
$N$      & 164  & Total respondents (82 per arm). \\
$\mu_t$  & 57   & Assumed mean support under treatment. \\
$\mu_c$  & 50   & Control mean, from the pilot. \\
$\sigma$ & 20   & Standard deviation, from the pilot. \\
$\alpha$ & 0.05 & Two-sided, by convention. \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.2em}

\small Assumed effect: $\mu_t - \mu_c = 7$ points, or $0.35\sigma$. \textbf{Can we detect a 7-point effect with 164 respondents?}


## Step 1: Do the Arithmetic

\vspace{0.2em}

\begin{align*}
\text{power} &= \Phi\!\left( \frac{|\mu_t - \mu_c|\sqrt{N}}{2\sigma} \;-\; \Phi^{-1}\!\left(1-\frac{\alpha}{2}\right) \right)
   && \text{\small the formula} \\[0.25em]
&= \Phi\!\left( \frac{7\,\sqrt{164}}{2(20)} \;-\; \textcolor{statsecondary}{1.96} \right)
   && \text{\small plug in; \;\; $\Phi^{-1}(0.975) = 1.96$} \\[0.25em]
&= \Phi\!\left( \textcolor{statsecondary}{\mathbf{2.24}} \;-\; 1.96 \right) \;=\; \textcolor{statsecondary}{\Phi(\mathbf{0.28})}
   && \text{\small $7 \times 12.81 / 40 = 2.24$}
\end{align*}

\vspace{0.5em}

\begin{center}
\small
$\underbrace{\mathbf{2.24}}_{\text{what you expect}}$ \qquad
$\underbrace{\mathbf{1.96}}_{\text{what you need}}$ \qquad
$\underbrace{\mathbf{0.28}}_{\text{all the room you have}}$
\end{center}

\vspace{0.5em}

\begin{center}
\scriptsize A thin margin means an unreliable study. One lookup turns $\Phi(0.28)$ into a number. \\
(Where does $1.96$ come from? Running the $z$-table \textit{backwards} --- see the appendix.)
\end{center}


## Step 2: Read the Answer Off the Table

\begin{columns}[T]

\begin{column}{0.30\textwidth}
\vspace{0.6em}

\begin{center}
$\text{power} = \Phi(\mathbf{0.28})$
\end{center}

\vspace{0.5em}

\footnotesize
Split the $z$-score across the two margins of the table:

\vspace{0.4em}

\hspace{0.5em}$\mathbf{0.2}\;$ selects the \textbf{row}

\vspace{0.2em}

\hspace{0.5em}$\mathbf{0.08}$ selects the \textbf{column}

\vspace{0.5em}

The cell where they meet is the area to the \textit{left} of $z$, which is exactly the power.
\end{column}

\begin{column}{0.70\textwidth}
\begin{center}
\scriptsize
\setlength{\tabcolsep}{2.4pt}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{c|cccccccccc}
\toprule
$z$ & .00 & .01 & .02 & .03 & .04 & .05 & .06 & .07 & \cellcolor{stataccbg}\textbf{.08} & .09 \\
\midrule
0.0 & .5000 & .5040 & .5080 & .5120 & .5160 & .5199 & .5239 & .5279 & \cellcolor{stataccbg}.5319 & .5359 \\
0.1 & .5398 & .5438 & .5478 & .5517 & .5557 & .5596 & .5636 & .5675 & \cellcolor{stataccbg}.5714 & .5753 \\
\cellcolor{stataccbg}\textbf{0.2} & \cellcolor{stataccbg}.5793 & \cellcolor{stataccbg}.5832 & \cellcolor{stataccbg}.5871 & \cellcolor{stataccbg}.5910 & \cellcolor{stataccbg}.5948 & \cellcolor{stataccbg}.5987 & \cellcolor{stataccbg}.6026 & \cellcolor{stataccbg}.6064 & \cellcolor{statorange}\textbf{\color{white}.6103} & \cellcolor{stataccbg}.6141 \\
0.3 & .6179 & .6217 & .6255 & .6293 & .6331 & .6368 & .6406 & .6443 & \cellcolor{stataccbg}.6480 & .6517 \\
0.4 & .6554 & .6591 & .6628 & .6664 & .6700 & .6736 & .6772 & .6808 & \cellcolor{stataccbg}.6844 & .6879 \\
0.5 & .6915 & .6950 & .6985 & .7019 & .7054 & .7088 & .7123 & .7157 & \cellcolor{stataccbg}.7190 & .7224 \\
\bottomrule
\end{tabular}
\end{center}
\end{column}

\end{columns}

\vspace{0.5em}

\begin{center}
\large $\text{power} \;=\; \Phi(0.28) \;=\; \textcolor{statorange}{\mathbf{0.61}}$
\end{center}


## The Smallest Effect This Study Can See

\small We now know what this design does with a 7-point effect. The next question is what it \textit{can} do: fix power at 80\% and solve the same formula for the effect instead of for power.

\vspace{0.3em}

\begin{center}
$\displaystyle \text{MDE} \;=\; \underbrace{\left(\Phi^{-1}\!\left(1-\tfrac{\alpha}{2}\right) + \Phi^{-1}(0.80)\right)}_{1.96 \;+\; 0.84 \;=\; 2.80} \;\times\; \underbrace{\frac{2\sigma}{\sqrt{N}}}_{\text{standard error}}$
\end{center}

\vspace{0.3em}

\begin{center}
\scriptsize
\setlength{\tabcolsep}{9pt}
\renewcommand{\arraystretch}{1.08}
\begin{tabular}{ccc}
\toprule
$N$ & \textbf{Smallest detectable effect} & \textbf{In standard deviations} \\
\midrule
\textcolor{statorange}{\textbf{164}} & \textcolor{statorange}{\textbf{8.75 points}} & \textcolor{statorange}{\textbf{0.44$\sigma$}} \\
257  & 6.99 & 0.35$\sigma$ \\
400  & 5.60 & 0.28$\sigma$ \\
800  & 3.96 & 0.20$\sigma$ \\
1396 & 3.00 & 0.15$\sigma$ \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.3em}

\begin{center}
\small Pilot assumes $\mu_t - \mu_c = \mathbf{7}$. \qquad This design detects $\textcolor{statorange}{\mathbf{8.75}}$ and up.
\end{center}


## Check: Put 8.75 Back Through the Formula

\vspace{0.2em}

\begin{align*}
\text{power} &= \Phi\!\left( \frac{8.75\,\sqrt{164}}{2(20)} \;-\; 1.96 \right)
   && \text{\small the MDE in place of the pilot's effect} \\[0.25em]
&= \Phi\!\left( \textcolor{statsecondary}{\mathbf{2.80}} \;-\; 1.96 \right) \;=\; \Phi(\textcolor{statsecondary}{\mathbf{0.84}})
   && \text{\small $8.75 \times 12.81 / 40 = 2.80$} \\[0.25em]
&= \mathbf{0.80}
   && \text{\small row $0.8$, column $.04 \to .7995$}
\end{align*}

\vspace{0.4em}

\begin{center}
\footnotesize
\setlength{\tabcolsep}{9pt}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{>{\raggedright\arraybackslash}p{6.6cm}c}
\toprule
\textbf{Pin these three} & \textbf{Get} \\
\midrule
$N = 164$, \; $\sigma = 20$, \; effect $= 7$      & power $= 0.61$ \\
$N = 164$, \; $\sigma = 20$, \; power $= 0.80$    & effect $= 8.75$ \\
\bottomrule
\end{tabular}
\end{center}



## Your Turn: Is the Tutoring Program Worth Studying?

\vspace{0.5em}

\begin{notebox}
\small A colleague is evaluating a new tutoring program. \textbf{400 students} will be randomly assigned, 200 to tutoring and 200 to business as usual, with final exam score as the outcome. Past cohorts averaged \textbf{72} points with a standard deviation of \textbf{10}, and the program is expected to add \textbf{3} points.
\end{notebox}

\vspace{0.7em}

\begin{center}
\large \textbf{What is the power of this design?}
\end{center}

\vspace{0.7em}

\begin{columns}[c]

\begin{column}{0.46\textwidth}
\begin{center}
\footnotesize
\setlength{\tabcolsep}{10pt}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{cc}
\toprule
\textbf{Symbol} & \textbf{Value} \\
\midrule
$N$      & 400  \\
$\mu_t$  & 75   \\
$\mu_c$  & 72   \\
$\sigma$ & 10   \\
$\alpha$ & 0.05, two-sided \\
\bottomrule
\end{tabular}
\end{center}
\end{column}

\begin{column}{0.50\textwidth}
\footnotesize
The formula and the $z$-table are on your \textbf{handout}.

\vspace{0.6em}

Work it in three moves:

\vspace{0.3em}

\hspace{0.4em}1. Compute what you \textbf{expect}

\vspace{0.2em}

\hspace{0.4em}2. Subtract what you \textbf{need}

\vspace{0.2em}

\hspace{0.4em}3. Look the difference up
\end{column}

\end{columns}



## Your Turn: The Solution

\vspace{0.2em}

\begin{align*}
\text{power} &= \Phi\!\left( \frac{|75 - 72|\,\sqrt{400}}{2(10)} \;-\; \Phi^{-1}\!\left(1-\tfrac{0.05}{2}\right) \right)
   && \text{\small \textbf{Step 1:} plug in} \\[0.25em]
&= \Phi\!\left( \frac{3 \times 20}{20} \;-\; 1.96 \right)
   && \text{\small \textbf{Step 2:} $\sqrt{400} = 20$; $\Phi^{-1}(0.975) = 1.96$} \\[0.25em]
&= \Phi(3.00 - 1.96) \;=\; \textcolor{statsecondary}{\Phi(\mathbf{1.04})}
   && \text{\small \textbf{Step 3:} simplify} \\[0.25em]
&= \mathbf{0.85}
   && \text{\small \textbf{Step 4:} row $1.0$, column $.04 \to .8508$}
\end{align*}

\vspace{0.2em}


## Check Your Work: The EGAP Power Calculator

\vspace{0.3em}

\begin{center}
\small \url{https://egap.shinyapps.io/power-app/}
\end{center}

\vspace{0.6em}

\begin{center}
\small
\setlength{\tabcolsep}{7pt}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{c>{\raggedright\arraybackslash}p{6.4cm}c}
\toprule
\textbf{Our symbol} & \textbf{Field in the app} & \textbf{Enter} \\
\midrule
$\mu_t - \mu_c$ & Treatment Effect Size                  & 3    \\
$\sigma$        & Standard Deviation of Outcome Variable & 10   \\
$\alpha$        & Significance Level                     & 0.05 \\
                & Power Target                           & 0.8  \\
                & Maximum Number of Subjects             & 800  \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.5em}

\begin{center}
\small Leave \textit{Clustered Design} and \textit{Binary Dependent Variable} unchecked.
\end{center}


## Reading the Power Curve

\begin{columns}[c]

\begin{column}{0.53\textwidth}
\includegraphics[width=\linewidth]{images/egap-power-app.png}
\end{column}

\begin{column}{0.45\textwidth}
\footnotesize
\setlength{\itemsep}{0.35em}
\begin{itemize}
\item The $x$-axis is \textbf{total sample size}; the $y$-axis is \textbf{power}.
\item The blue curve traces power as $N$ grows, holding the effect at 3 and $\sigma$ at 10.
\item The red dashed line marks the \textbf{80\% target}. Where the curve crosses it is the $N$ you need (here \textbf{349}).
\item Our $N = 400$ sits to the \textit{right} of that crossing. Reading up from 400 gives $\approx \mathbf{0.85}$.
\item The curve \textbf{flattens}: past $\sim$600 suggests diminishing returns at larger N.
\end{itemize}
\end{column}

\end{columns}

## Where to Situate Power Analysis in Your Research

- Power analysis is tricky because we need information about our data to derive power (e.g., $\sigma$, $\mu_t - \mu_c$).
- The problem is power calculations are most useful *before* we've collected our data!
- In order to do power analysis we need to do some educated guessing: 
  - How big an effect should I anticipate? 
  - How many people can I realistically afford to treat?
  - How many subjects will answer my survey?

## Conjecture in Power Analysis

If power analysis relies on educated guesses, what can we build those guesses from?

\vspace{0.4em}

\begin{center}
\footnotesize
\setlength{\tabcolsep}{7pt}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{>{\raggedright\arraybackslash}p{4.0cm}>{\raggedright\arraybackslash}p{8.7cm}}
\toprule
\textbf{The guess} & \textbf{What you can build it from} \\
\midrule
How big an effect should I anticipate? \newline \textcolor{statsecondary}{$\mu_t - \mu_c$, \; $\sigma$}
& \leavevmode\uncover<2->{\textbullet~Published effects, discounted for publication bias}\newline
\uncover<2->{\textbullet~A pilot or baseline wave: gives you $\sigma$ directly}\newline
\uncover<2->{\textbullet~The \textit{smallest} effect big enough to matter} \\
\midrule
How many people can I realistically afford to treat? \newline \textcolor{statsecondary}{$N$}
& \leavevmode\uncover<3->{\textbullet~Budget divided by cost per subject}\newline
\uncover<3->{\textbullet~Field period, staff time, partner capacity}\newline
\uncover<3->{\textbullet~Size of the sampling frame you can actually reach} \\
\midrule
How many subjects will answer my survey? \newline \textcolor{statsecondary}{$N$ realized}
& \leavevmode\uncover<4->{\textbullet~Response rates from earlier waves or comparable surveys}\newline
\uncover<4->{\textbullet~Attrition rates in similar designs}\newline
\uncover<4->{\textbullet~Power the \textit{analyzed} $N$, not the invited $N$} \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.2em}

\uncover<5->{\begin{center}
\small Every one of these is a \textbf{range}, not a number. \textit{So which number do you use?}
\end{center}}


## What If We Don't Know the Effect?

\vspace{0.2em}

\small Back to the survey experiment. The pilot fixed everything \textit{except} for the anticipated effect size.

\vspace{0.5em}

\begin{columns}[T]

\begin{column}{0.40\textwidth}
\begin{center}
\footnotesize
\setlength{\tabcolsep}{6pt}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{cc}
\toprule
\multicolumn{2}{c}{\textbf{Fixed by the pilot}} \\
\midrule
$\mu_c$        & 50   \\
$\sigma$       & 20   \\
$\alpha$       & 0.05 \\
target power & 0.80 \\
\midrule
$\mu_t - \mu_c$ & \textcolor{statorange}{\textbf{?}} \\
$N$             & \textcolor{statorange}{\textbf{?}} \\
\bottomrule
\end{tabular}
\end{center}
\end{column}

\begin{column}{0.58\textwidth}
\begin{center}
\footnotesize
\setlength{\tabcolsep}{6pt}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{>{\raggedright\arraybackslash}p{5.6cm}c}
\toprule
\textbf{Where an effect could come from} & $\mu_t - \mu_c$ \\
\midrule
Our own pilot                          & 7 \\
A published study of a similar message & 5 \\
That published effect, discounted for publication bias & 3 \\
\bottomrule
\end{tabular}
\end{center}
\end{column}

\end{columns}

\vspace{0.5em}


## The Formula, in Code

```{=latex}
\vspace{0.6em}

\begin{center}
$\displaystyle \text{power} \;=\; \Phi\!\left( \frac{|\mu_t - \mu_c|\sqrt{N}}{2\sigma} \;-\; \Phi^{-1}\!\left(1-\frac{\alpha}{2}\right) \right)$
\end{center}

\vspace{0.9em}
```

```{r power-fn, echo=TRUE}
power_calc <- function(mu_t, mu_c, sigma, N, alpha = 0.05) {
  z  <- abs(mu_t - mu_c) * sqrt(N) / (2 * sigma)
  cv <- qnorm(1 - alpha / 2)
  pnorm(z - cv) + pnorm(-z - cv)     # both rejection tails
}
```

```{=latex}
\vspace{0.9em}

\begin{center}
\large $\Phi \;=\;$ \texttt{pnorm} \qquad\qquad $\Phi^{-1} \;=\;$ \texttt{qnorm}
\end{center}
```


## Power Across a Range of $N$

```{r sweep, echo=TRUE, message=FALSE, warning=FALSE}
grid <- expand.grid(N     = seq(60, 1600, by = 10),
                    delta = c(3, 5, 7))

grid$power <- power_calc(mu_t  = 50 + grid$delta,
                         mu_c  = 50,
                         sigma = 20,
                         N     = grid$N)

head(grid, 3)
```

```{=latex}
\vspace{0.4em}

\begin{center}
\small One formula, three assumptions, `r nrow(grid)` rows. Now plot \texttt{power} against \texttt{N}.
\end{center}
```


## How Many Respondents Do We Need?

```{r sweep-plot-setup, echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
grid$delta <- factor(grid$delta, levels = c(3, 5, 7))
nstar <- data.frame(delta = factor(c(3, 5, 7), levels = c(3, 5, 7)))
nstar$N <- ceiling((2 * 20 * (qnorm(0.975) + qnorm(0.80)) / c(3, 5, 7))^2)
nstar$power <- 0.80
ramp <- c("3" = "#4A90C2", "5" = "#1F5F8B", "7" = "#0B2E52")
```

```{r sweep-plot, echo=FALSE, fig.width=7.4, fig.height=3.3, dev="png", dpi=300, fig.align="center"}
ggplot(grid, aes(N, power, colour = delta)) +
  geom_hline(yintercept = 0.80, linetype = "dashed",
             colour = "#B41E1E", linewidth = 0.4) +
  geom_line(linewidth = 0.85) +
  geom_point(data = nstar, size = 1.9, colour = "#B41E1E") +
  geom_text(data = nstar, aes(label = N), vjust = 1.9, hjust = -0.15,
            size = 3, colour = "#B41E1E", show.legend = FALSE) +
  scale_colour_manual(values = ramp, name = "Assumed effect (points):") +
  scale_y_continuous(limits = c(0, 1.05), breaks = seq(0, 1, 0.2)) +
  scale_x_continuous(breaks = seq(0, 1600, 200)) +
  labs(x = "Total respondents (N)", y = "Power") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.25, colour = "grey88"),
        axis.title = element_text(colour = "grey30"),
        legend.position = "bottom", legend.margin = margin(t = -4),
        legend.key.width = unit(16, "pt"),
        legend.title = element_text(colour = "grey30"))
```

\vspace{0.1em}



## Your Turn: Build the Sweep in R

\small Your department is considering a six-week **peer-mentoring program** for first-year students, and can randomize students into it or onto a waitlist. The outcome is end-of-year GPA. The registrar tells you first-year GPA averages **3.05**, with a standard deviation of **0.45**.

\vspace{0.5em}

::: {.columns}
:::: {.column width="44%"}

```{=latex}
\begin{center}
\footnotesize
\setlength{\tabcolsep}{6pt}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{cc}
\toprule
\multicolumn{2}{c}{\textbf{Given}} \\
\midrule
$\mu_c$  & 3.05 \\
$\sigma$ & 0.45 \\
$\alpha$ & 0.05 \\
\bottomrule
\end{tabular}
\end{center}
```

::::
:::: {.column width="54%"}

```{=latex}
\vspace{0.4em}
\footnotesize
\textbf{You decide:}

\vspace{0.3em}

\hspace{0.4em}\textbullet~Which GPA gains are plausible?

\vspace{0.2em}

\hspace{0.4em}\textbullet~How many students could you enroll?
```

::::
:::

\vspace{0.4em}


## Your Turn: One Solution

```{r exercise-solution, echo=TRUE}
grid <- expand.grid(N     = seq(50, 1000, by = 10),
                    delta = c(0.10, 0.15, 0.20))
grid$power <- power_calc(mu_t = 3.05 + grid$delta, mu_c = 3.05,
                         sigma = 0.45, N = grid$N)

for (d in unique(grid$delta)) {
  ok <- grid$N[grid$delta == d & grid$power >= 0.80]
  cat("effect", d, "GPA points -> N =", min(ok), "\n")
}
```

\footnotesize Your effects and $N$ range will differ. The *structure* is the answer: one grid, one call, one threshold.


## Where the Formula Runs Out

\small The closed form is not general. It is designed for difference of means testing; if we want to use a more complex model, we need a different approach:

\vspace{0.4em}

\begin{center}
\small
\setlength{\tabcolsep}{7pt}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{>{\raggedright\arraybackslash}p{5.3cm}>{\raggedright\arraybackslash}p{7.4cm}}
\toprule
\textbf{It assumes} & \textbf{So it cannot handle} \\
\midrule
Two arms, split 50/50            & Three arms, unequal allocation \\
One continuous, normal outcome   & Binary, counts, skewed scales \\
A simple difference in means     & Covariates, fixed effects, IV, mixed models \\
Independent observations         & Clustered or repeated measures \\
Everyone recruited is analysed   & Attrition, non-response \\
\bottomrule
\end{tabular}
\end{center}



## Power with Monte Carlo Simulation

\vspace{0.6em}

\small
\begin{enumerate}
\item \textbf{Invent a world} where your assumptions are literally true: $\mu_t - \mu_c = 7$, $\sigma = 20$, $N = 164$.

\vspace{0.35em}

\item \textbf{Draw one fake study} from that world and run \textit{the analysis you actually plan to run}.

\vspace{0.35em}

\item \textbf{Repeat} a few thousand times. Power is the share of those studies where $p < 0.05$.
\end{enumerate}


## What Simulation Adds to the Sweep

\vspace{0.3em}

\begin{center}
\small
\setlength{\tabcolsep}{6pt}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{>{\raggedright\arraybackslash}p{2.7cm}>{\raggedright\arraybackslash}p{4.5cm}>{\raggedright\arraybackslash}p{5.6cm}}
\toprule
& \textbf{Sweep} & \textbf{Simulation} \\
\midrule
What varies      & Inputs to one formula & Inputs, \textit{and} the design and estimator \\
Power comes from & A closed-form CDF & The share of fake studies that reject \\
Cost             & Exact, instant & Approximate, seconds to minutes \\
Works for        & Two arms, means, complete data & Anything you can write code for \\
You also get     & One number & The whole distribution of estimates and $p$-values \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.4em}

\begin{center}
\small They are not rivals. The sweep is the \textbf{outer loop}; simulation changes what fills each cell.
\end{center}


## Estimand and Estimator

\vspace{0.4em}

\begin{center}
\small
\setlength{\tabcolsep}{7pt}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{>{\raggedright\arraybackslash}p{2.2cm}>{\raggedright\arraybackslash}p{5.0cm}>{\raggedright\arraybackslash}p{5.6cm}}
\toprule
& \textbf{What it is} & \textbf{Examples} \\
\midrule
\textbf{Estimand}  & The quantity you want to learn. A property of the \textit{world}. & The ATE; the effect among women; an interaction; a LATE \\
\textbf{Estimator} & The procedure you run on data. A property of your \textit{code}. & Difference in means; OLS with covariates; 2SLS; a mixed model \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.5em}



## Anatomy of One Simulation

\small Before automating anything, run the study **once**.

```{r one-draw, echo=TRUE}
set.seed(3)
one_study <- function() {
  z <- sample(rep(0:1, each = 82))             # RANDOMLY assign the message
  y <- rnorm(164, mean = 50 + 7 * z, sd = 20)  # ESTIMAND: the truth is 7
  coef(summary(lm(y ~ z)))["z", 4]             # ESTIMATOR: what you run
}

c(study_1 = one_study(), study_2 = one_study(), study_3 = one_study())
```

\begin{center}
\small \textbf{Power is just the share of these that land under 0.05.}
\end{center}


## How Many Simulations Is Enough?

\small A simulated power estimate is itself an estimate. Its standard error is $\sqrt{\text{power}(1-\text{power})/S}$.

\vspace{0.5em}

\begin{center}
\small
\setlength{\tabcolsep}{9pt}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{cc>{\raggedright\arraybackslash}p{6.4cm}}
\toprule
\textbf{$S$} & \textbf{$\pm$ at power $= 0.80$} & \textbf{Good for} \\
\midrule
200    & 0.028 & A first look, nothing more \\
1{,}000  & 0.013 & Exploring designs \\
2{,}000  & 0.009 & Most teaching and prototyping \\
10{,}000 & 0.004 & A number going in a grant or pre-registration \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.5em}

\begin{center}
\small The cheap check: rerun with a different seed. If the answer moves more than you can live with, raise $S$.
\end{center}


## Building a Monte Carlo simulation for Power

\small Keep the sweep; replace what fills the column. Run the study thousands of times and count.

```{r mc-harness, echo=TRUE}
sim_power <- function(N, tau, sigma, S = 2000) {
  mean(replicate(S, {
    z <- sample(rep(0:1, length.out = N))
    y <- rnorm(N, 50 + tau * z, sigma)  # the ESTIMAND
    coef(summary(lm(y ~ z)))["z", 4]    # the ESTIMATOR
  }) < 0.05)
}
set.seed(11)
c(simulation = sim_power(164, 7, 20),
  formula    = power_calc(57, 50, 20, 164))
```


## What If I Change My Analysis Plan?

\small In the survey experiment you planned to compare mean thermometer scores. Now you decide to fit a regression that adjusts for each respondent's *baseline* thermometer reading, taken before the message.

\vspace{0.5em}

\begin{center}
\footnotesize
\setlength{\tabcolsep}{7pt}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{>{\raggedright\arraybackslash}p{2.7cm}>{\raggedright\arraybackslash}p{4.8cm}>{\raggedright\arraybackslash}p{5.2cm}}
\toprule
& \textbf{What you planned} & \textbf{What you now plan} \\
\midrule
The code & \texttt{t.test(y \textasciitilde\ z)} & \texttt{lm(y \textasciitilde\ z + x)} \\
Estimand & The ATE of the message & \textcolor{highlightgreen}{\textbf{the same ATE}} \\
Estimator & Difference in means & OLS with a covariate \\
Does the closed form describe it? & Yes & \textcolor{errorred}{\textbf{No}} \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.3em}

\begin{center}
\small You changed \textbf{how you will answer}, not \textbf{what you are asking}.
\end{center}

\vspace{0.2em}

\scriptsize The baseline reading `x` is measured before treatment, so it is a *precision covariate*, not a confounder; it correlates about 0.6 with the outcome. Does adjusting help or hurt? **Simulate it.**


## Branch 1: Same Estimand, Better Estimator

```{r mc-covariate, echo=TRUE}
sim_cov <- function(adjust, N = 164, S = 2000, tau = 7) {
  mean(replicate(S, {
    z <- sample(rep(0:1, length.out = N)); x <- rnorm(N)  # x = baseline
    y <- 50 + tau * z + 12 * x + rnorm(N, 0, 16) # cor(x, y) = 0.6
    fit <- if (adjust) lm(y ~ z + x) else lm(y ~ z)
    coef(summary(fit))["z", 4]
  }) < 0.05)
}
set.seed(11)
c(unadjusted = sim_cov(FALSE), adjusted = sim_cov(TRUE))
```

\begin{center}
\small \textbf{80\%} power at the same $N$: adjusting drops $\sigma$ from 20 to 16, at \textbf{zero} extra cost in respondents.
\end{center}


## The Smallest Effect the Regression Can See

\small The formula does not describe `lm(y ~ z + x)`, so fix power at 80\% and sweep the **effect** by simulation, holding $N$ at 164.

```{r mde-sim, echo=TRUE}
set.seed(5)
taus <- 5:9                                    # candidate effects, in points
pw   <- sapply(taus, function(t) sim_cov(TRUE, S = 1000, tau = t))
mde <- function(p, tau, S = 1000, target = 0.80) {   # fit, then invert
  b <- coef(glm(cbind(round(p*S), S - round(p*S)) ~ log(tau), binomial))
  exp((qlogis(target) - b[[1]]) / b[[2]])
}
c(setNames(round(pw, 2), taus), MDE = round(mde(pw, taus), 2))
```

\footnotesize Adjusting moves the MDE from \textbf{8.75} points to \textbf{6.9}. The closed form with $\sigma = 16$ gives 7.0.


## What If x Is a Confounder Instead?

\scriptsize Same covariate, but `z` is now **not randomized**: higher `x` makes treatment likelier. Watch the rejection rate.

```{r mc-confound, echo=TRUE}
set.seed(3)
sim_obs <- function(adjust, N = 164, S = 2000) {
  out <- replicate(S, {
    x <- rnorm(N); z <- rbinom(N, 1, plogis(1.2 * x))  # z DEPENDS on x
    y <- 50 + 7 * z + 12 * x + rnorm(N, 0, 16)         # truth is still 7
    fit <- if (adjust) lm(y ~ z + x) else lm(y ~ z)
    c(coef(fit)["z"], coef(summary(fit))["z", 4]) })
  c(estimate = mean(out[1, ]), rejection = mean(out[2, ] < 0.05)) }
obs <- rbind(`y ~ z` = sim_obs(FALSE), `y ~ z + x` = sim_obs(TRUE)); obs
```

\scriptsize The unadjusted model rejects \textbf{100\%} of the time. That is \textit{not} power --- the estimate is centred on 18.2 when the truth is 7. A rejection rate is only power once the estimator is unbiased.



## Branch 2: Same Study, Different Estimand

\small The message moves $g=1$ by 10 points and $g=0$ by 4. Average effect still 7; the \textbf{interaction} is 6.

```{r mc-interaction, echo=TRUE}
set.seed(11)
p <- replicate(2000, {
  z <- sample(rep(0:1, each = 82)); g <- rep(0:1, 82)
  y <- rnorm(164, 50 + 4 * z + 6 * z * g + 2 * g, 20)
  c(ate = coef(summary(lm(y ~ z)))["z", 4],
    int = coef(summary(lm(y ~ z * g)))["z:g", 4])
})
rowMeans(p < 0.05)
```

\begin{center}
\small Nothing changed but the \textbf{question}. An interaction carries twice the standard error, so it behaves like a main effect with $\sigma$ \textbf{doubled}: 80\% power needs $\approx$\textbf{1400} respondents, against 257 for the ATE.
\end{center}


## Power Is a Property of the Triple

\vspace{0.4em}

\begin{center}
\small
\setlength{\tabcolsep}{7pt}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{>{\raggedright\arraybackslash}p{2.4cm}>{\raggedright\arraybackslash}p{5.6cm}>{\raggedright\arraybackslash}p{4.5cm}}
\toprule
\textbf{Choice} & \textbf{What it is} & \textbf{What it did here} \\
\midrule
\textbf{Estimand}  & The quantity you are after: ATE, interaction, subgroup, LATE & $0.61 \rightarrow 0.15$ \\
\textbf{Estimator} & The model you will actually fit & $0.61 \rightarrow 0.80$ \\
\textbf{Design}    & $N$, allocation, blocking, what gets randomized & the lever fully in your hands \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.6em}

\begin{center}
\small Power is not a property of your sample size. It is a property of all three together. \\[0.4em]
\textbf{What is my estimand, what estimator will I actually run, and what does \textit{that pair} need?}
\end{center}


## What Voids Your Calculation

\small Everything so far assumed you randomize individuals, one at a time, on a continuous outcome, and analyse everyone you recruited. Break any of those and the number you just computed is wrong --- usually by a lot.

\vspace{0.3em}

\begin{center}
\scriptsize
\setlength{\tabcolsep}{6pt}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{>{\raggedright\arraybackslash}p{4.0cm}>{\raggedright\arraybackslash}p{3.2cm}>{\raggedright\arraybackslash}p{5.3cm}}
\toprule
\textbf{If your design has\ldots} & \textbf{Your $N$ is wrong by} & \textbf{What that looked like here} \\
\midrule
Groups randomized, not individuals & $\times\;\text{DEFF} = 1 + (m-1)\rho$
  & 283 $\rightarrow$ \textcolor{errorred}{\textbf{1{,}302}} students; analysed naively, a \textcolor{errorred}{\textbf{36\%}} false-positive rate \\
Take-up below 100\%    & $\div$ take-up$^2$      & 50\% take-up $\rightarrow$ \textbf{4$\times$} the sample: 257 $\rightarrow$ 1{,}028 \\
Attrition              & $\div\;(1 - \text{loss})$ & 20\% loss $\rightarrow$ recruit \textbf{322} to analyse 257 \\
A binary outcome       & $\sigma = \sqrt{p(1-p)}$  & The same 164 people: 61\% power continuous, \textbf{36\%} for 30\%$\rightarrow$42\% \\
A subgroup or interaction & $\approx\;4\times$     & 257 $\rightarrow$ \textbf{$\sim$1{,}400} \\
\bottomrule
\end{tabular}
\end{center}


## Your Turn: Simulate a Design of Your Own

\small Take a study you are actually planning, or one you have just read. Before writing any code, answer three questions:

\vspace{0.6em}

\small
\begin{enumerate}
\item \textbf{What is my estimand?} An average effect, a difference between groups, an effect for compliers only?

\vspace{0.3em}

\item \textbf{What estimator will I actually run?} A difference in means, OLS with controls, something with clustered errors?

\vspace{0.3em}

\item \textbf{What must I assume to simulate it?} A baseline mean, an outcome standard deviation, a plausible effect.
\end{enumerate}

\vspace{0.7em}

\begin{center}
\small Then fill in the scaffold on the next slide and sweep it over $N$.
\end{center}


## Your Turn: A Scaffold

```{r scaffold, eval=FALSE, echo=TRUE}
one_study <- function(N) {
  z <- sample(rep(0:1, length.out = N))  # 1. ASSIGN: randomize treatment

  y <- rnorm(N, mean = ???, sd = ???) # 2. GENERATE: your ESTIMAND

  coef(summary(lm(y ~ z)))["z", 4]    # 3. ANALYSE: your ESTIMATOR
}

power_for <- function(N, S = 1000) mean(replicate(S, one_study(N)) < 0.05)

sapply(seq(100, 600, by = 50), power_for)
```

\small Change line 2 and you have changed your estimand. Change line 3 and you have changed your estimator. Everything else stays put.


## Appendix: Where 1.96 Comes From

\begin{columns}[T]

\begin{column}{0.30\textwidth}
\vspace{0.6em}

\begin{center}
$\Phi^{-1}(\mathbf{0.975})$
\end{center}

\vspace{0.5em}

\footnotesize
$\Phi^{-1}$ runs the table \textbf{backwards}. You are handed an area and must find the $z$ that produces it:

\vspace{0.4em}

\hspace{0.5em}1. Hunt $.9750$ in the \textbf{body}

\vspace{0.2em}

\hspace{0.5em}2. Read off its \textbf{margins}

\vspace{0.5em}

Row $1.9$ plus column $.06$ gives the critical value you already know: $\mathbf{1.96}$.
\end{column}

\begin{column}{0.70\textwidth}
\begin{center}
\scriptsize
\setlength{\tabcolsep}{2.4pt}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{c|cccccccccc}
\toprule
$z$ & .00 & .01 & .02 & .03 & .04 & .05 & \cellcolor{stataccbg}\textbf{.06} & .07 & .08 & .09 \\
\midrule
1.6 & .9452 & .9463 & .9474 & .9484 & .9495 & .9505 & \cellcolor{stataccbg}.9515 & .9525 & .9535 & .9545 \\
1.7 & .9554 & .9564 & .9573 & .9582 & .9591 & .9599 & \cellcolor{stataccbg}.9608 & .9616 & .9625 & .9633 \\
1.8 & .9641 & .9649 & .9656 & .9664 & .9671 & .9678 & \cellcolor{stataccbg}.9686 & .9693 & .9699 & .9706 \\
\cellcolor{stataccbg}\textbf{1.9} & \cellcolor{stataccbg}.9713 & \cellcolor{stataccbg}.9719 & \cellcolor{stataccbg}.9726 & \cellcolor{stataccbg}.9732 & \cellcolor{stataccbg}.9738 & \cellcolor{stataccbg}.9744 & \cellcolor{statorange}\textbf{\color{white}.9750} & \cellcolor{stataccbg}.9756 & \cellcolor{stataccbg}.9761 & \cellcolor{stataccbg}.9767 \\
2.0 & .9772 & .9778 & .9783 & .9788 & .9793 & .9798 & \cellcolor{stataccbg}.9803 & .9808 & .9812 & .9817 \\
2.1 & .9821 & .9826 & .9830 & .9834 & .9838 & .9842 & \cellcolor{stataccbg}.9846 & .9850 & .9854 & .9857 \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.6em}

\small Start \textit{inside} the table and read outward: the mirror image of Step 2.
\end{column}

\end{columns}

\vspace{0.4em}

\begin{center}
$\Phi^{-1}(0.975) = \mathbf{1.96}$: the bar the study has to clear.
\end{center}


## Appendix: Two Loops, Not One

\small The sweep and the simulation stack. Neither replaces the other.

\vspace{0.4em}

\begin{center}
\small
\setlength{\tabcolsep}{7pt}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{>{\raggedright\arraybackslash}p{2.6cm}>{\raggedright\arraybackslash}p{4.4cm}>{\raggedright\arraybackslash}p{5.6cm}}
\toprule
& \textbf{What it loops over} & \textbf{Question it answers} \\
\midrule
\textbf{Outer} \newline \texttt{expand.grid} & The designs you are considering: each $N$, each estimator & ``Which study should I run?'' \\
\textbf{Inner} \newline \texttt{replicate}   & Random draws from one fixed design & ``How often does \textit{this} study win?'' \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.4em}

```{r nesting, eval=FALSE, echo=TRUE}
grid <- expand.grid(N = seq(120, 300, by = 30),      # OUTER
                    adjust = c(FALSE, TRUE))
grid$power <- mapply(function(n, a) sim_cov(a, n),   # INNER runs inside
                     grid$N, grid$adjust)
```


## Appendix: Sweeping the Simulation

```{r sim-sweep, echo=TRUE}
set.seed(4); Ns <- seq(120, 300, by = 30)
sweep_sim <- data.frame(N = Ns,
  unadjusted = sapply(Ns, function(n) sim_cov(FALSE, n, S = 1500)),
  adjusted   = sapply(Ns, function(n) sim_cov(TRUE,  n, S = 1500)))
sweep_sim
```

\scriptsize Unadjusted crosses 80\% between 240 and 270; adjusted, between 150 and 180. The closed form puts the two at **257** and **165**.


## Appendix: Finding the Cutoff by Simulation

```{r sim-plot, echo=FALSE, fig.width=7.6, fig.height=1.10, dev="png", dpi=300, fig.align="center"}
library(ggplot2)
long <- data.frame(
  N         = rep(sweep_sim$N, 2),
  power     = c(sweep_sim$unadjusted, sweep_sim$adjusted),
  estimator = rep(c("y ~ z", "y ~ z + x"), each = nrow(sweep_sim)))

ggplot(long, aes(N, power, colour = estimator)) +
  geom_hline(yintercept = 0.80, linetype = "dashed", colour = "#B41E1E",
             linewidth = 0.4) +
  geom_line(linewidth = 0.8) + geom_point(size = 1.7) +
  scale_colour_manual(values = c("y ~ z" = "#4A90C2", "y ~ z + x" = "#0B2E52"),
                      name = NULL) +
  scale_y_continuous(limits = c(0.4, 1), breaks = seq(0.4, 1, 0.2)) +
  labs(x = "Total respondents (N)", y = "Simulated power") +
  theme_minimal(base_size = 8) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.25, colour = "grey88"),
        axis.title = element_text(colour = "grey30"),
        legend.position = "right", legend.key.width = unit(14, "pt"),
        plot.margin = margin(2, 2, 2, 2))
```

```{r cutoff-fit, echo=TRUE}
cutoff <- function(p, N, S = 1500, target = 0.80) {   # fit, then invert
  b <- coef(glm(cbind(round(p*S), S - round(p*S)) ~ log(N), binomial))
  exp((qlogis(target) - b[1]) / b[2])
}
round(c(adjusted   = cutoff(sweep_sim$adjusted,   sweep_sim$N),
        unadjusted = cutoff(sweep_sim$unadjusted, sweep_sim$N)))
```

\scriptsize Do **not** run `approx()` over simulated power: noise makes the curve non-monotone. Fit a monotone curve, invert that.


## Appendix: The Same Covariate, Two Different Jobs

\small Which job `x` is doing depends entirely on how `z` was assigned.

\vspace{0.5em}

\begin{center}
\footnotesize
\setlength{\tabcolsep}{6pt}
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{>{\raggedright\arraybackslash}p{3.0cm}>{\raggedright\arraybackslash}p{4.7cm}>{\raggedright\arraybackslash}p{4.9cm}}
\toprule
& \textbf{Randomized:} $x$ is a \textit{precision covariate} & \textbf{Observational:} $x$ is a \textit{confounder} \\
\midrule
Why include it & Optional. Randomization already gives the right answer. & Mandatory. Without it the answer is wrong. \\
Estimate, \texttt{y \textasciitilde\ z} & 7.00 \; \footnotesize(truth is 7) & \textcolor{errorred}{\textbf{`r round(obs["y ~ z", "estimate"], 2)`}} \; \footnotesize(truth is 7) \\
Adding $x$ & Soaks up outcome variance $\Rightarrow$ \textcolor{highlightgreen}{\textbf{more power}} & Also soaks up \textit{treatment} variance $\Rightarrow$ \textcolor{errorred}{\textbf{less power}} \\
Power, \texttt{y \textasciitilde\ z + x} & \textbf{0.80} & `r round(obs["y ~ z + x", "rejection"], 2)` \\
\bottomrule
\end{tabular}
\end{center}



## Appendix: When You Randomize Groups, Not People

\small Assign whole classrooms, clinics, or villages and observations within a cluster are correlated. Effective sample size shrinks by the \textbf{design effect}:

\vspace{0.3em}

\begin{center}
$\text{DEFF} \;=\; 1 + (m - 1)\,\rho$ \qquad \footnotesize $m$ = cluster size, \; $\rho$ = intra-cluster correlation
\end{center}

\vspace{0.5em}

\begin{center}
\small
\setlength{\tabcolsep}{7pt}
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{>{\raggedright\arraybackslash}p{7.2cm}c}
\toprule
\textbf{GPA study,} $\delta = 0.15$ & \textbf{Students needed} \\
\midrule
Randomize individual students                        & 283 \\
Randomize classrooms of 25, $\rho = 0.15$ (DEFF 4.6) & \textbf{1300} \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.5em}

\begin{center}
\small That checkbox we told you to leave unchecked in the EGAP app? \textit{This is what it does.}
\end{center}


## Appendix: Clustering in Simulation

\small The danger: the naive analysis still returns a number, and the number looks fine.

```{r mc-cluster, echo=TRUE}
set.seed(7)
n_clus <- 12; m <- 25                  # 300 students in 12 classrooms
sb <- 0.45 * sqrt(0.15); sw <- 0.45 * sqrt(0.85)
p <- replicate(1500, {
  zc <- sample(rep(0:1, each = n_clus / 2))  # assign CLASSROOMS, not students
  u  <- rnorm(n_clus, 0, sb)           # classroom-level shock
  y  <- rnorm(n_clus * m, rep(3.05 + 0.15 * zc + u, each = m), sw)
  id <- rep(1:n_clus, each = m)
  c(as_individuals = t.test(y ~ rep(zc, each = m))$p.value,
    as_clusters    = t.test(tapply(y, id, mean) ~ zc)$p.value)
})
rowMeans(p < 0.05)
```
